Appendix 3: Estimating species interactions

### 2015-08-04 Harris_2015/manuscript-materials/draft.Rmd
other things
almost done

@@

If nontrophic species interactions, such as competition, are important drivers
of community assembly, then ecologists might expect to see their influence in
species composition data [@macarthur_population_1958; @diamond_island_1975].
Despite decades of work and several major controversies, however
[@lewin_santa_1983; @strong_ecological_1984; @gotelli_swap_2003;
@connor_checkered_2013], existing methods for detecting competition's effects
on community structure are unreliable [@gotelli_empirical_2009]. In particular,
species' effects on one another can become lost in the complex web of direct and
indirect interactions in real assemblages. For example, the competitive
interaction between the two shrub species in Figure 1A can become obscured by
their shared tendency to occur in unshaded areas (Figure 1B). If this sort of
effect is common, then the vast majority of ecologists' methods, which rely on
test statistics describing the total number of co-occurrences or overall
correlation between putatively interacting species [@diamond_island_1975;
@strong_ecological_1984; @gotelli_empirical_2009; @veech_probabilistic_2013;
@pollock_understanding_2014] will not generally provide much evidence regarding
species' direct effects on one another. The goal of this paper is to resolve
this long-standing problem in community ecology.

@@

In this paper, following recent work by @azaele_inferring_2010 and
@fort_statistical_2013, I show that Markov networks [undirected graphical models
also known as Markov random fields; @murphy_machine_2012] can provide a
framework for understanding the landscape-level consequences of pairwise species
interactions, and  for detecting them from observed presence-absence matrices.
Markov networks have been used in many scientific fields for decades in similar
contexts, from physics [where nearby particles interact magnetically with one
another; @cipra_introduction_1987] to spatial statistics [where adjacent grid
cells interact with one another; @harris_contact_1974; @gelfand_modelling_2005].
While community ecologists explored some related approaches in the 1980's
[@whittam_species_1981], computational limitations had previously imposed severe
approximations that led to unintelligible results [e.g. "probabilities" greater
than one; @gilpin_factors_1982].

The rest of the paper proceeds as follows. First, I discuss how Markov networks
operate and how they can be used to simulate landscape-level data or to make
exact predictions regarding the direct and indirect consequences of possible
interaction matrices. Then, using simulated data sets where the "true"
ecological structure is known, I compare this approach with several existing
methods for detecting species interactions. Finally, I discuss opportunities for
extending the approach presented here to other problems in community ecology,
such as disentangling the effects of biotic and abiotic interactions on species
composition.

@@

#### Markov networks.
Markov networks provide a framework for translating back and forth between the
conditional relationships among species (Figure 1C) and the kinds of species
assemblages that these relationships produce. In this section, I show how a set
of conditional relationships can be used to determine the how groups of species
can co-occur. Methods for estimating conditional relationships from
presence-absence data are discussed in the next section.

@@

Here, $\alpha_{i}$ is an intercept term determining the amount that the presence
of species $i$ contributes to the log-probability of $\vec{y}$; it directly
controls the prevalence of species $i$. Similarly, $\beta_{ij}$ is the amount
that the co-occurrence of species $i$ and species $j$ contributes to the
log-probability; it controls the conditional relationship between two species,
i.e. the probability that they will be found together, after controlling for the
other species in the network (Figure 2A, Figure 2B). For example, $\beta_{ij}$
might have a value of $+2$ for two mutualists, indicating that the odds of
observing one species are $e^2$ times higher its partner is present than when
it is absent. Because the relative probability of a presence-absence vector
increases when positively-associated species co-occur and decreases when
negatively-associated species co-occur, the model tends---all else equal---to
produce assemblages that have many pairs of positively-associated species and
relatively few pairs of negatively-associated species (exactly as an ecologist
might expect).

@@

#### Estimating $\alpha$ and $\beta$ coefficients from presence-absence data.
In the previous section, the values of $\alpha$ and $\beta$ were known and the
goal was to make predictions about possible species assemblages. In
practice, however, ecologists will often need to estimate the parameters from
an observed co-occurrence matrix (i.e. from a matrix of ones and zeros
indicating which species are present at which sites). When the number of species
is reasonably small, one can compute exact maximum likelihood estimates for all
of the $\alpha$ and $\beta$ coefficients given a presence-absence matrix by
optimizing $p(\vec{y}; \alpha, \beta)$. Fully-observed Markov networks like the
ones considered here have unimodal likelihood surfaces [@murphy_machine_2012],
ensuring that this procedure will always converge on the global maximum. This
maximum represents the unique combination of $\alpha$ and $\beta$ coefficients
that would be expected to produce exactly the observed co-occurrence frequencies
on average (i.e. it matches the sufficient statistics of the model distribution
to the sufficient statistics of the data). For the analyses in this paper, I
used the *rosalia* package [@harris_rosalia_2015] for the R programming language
[@r_core_team_r_2015] to optimize the Markov network parameters.

#### Simulated landscapes.
In order to compare different methods, I simulated two sets of landscapes using
known parameters.

The first set of simulated landscapes included the three competing species
shown in Figure 1. For each of 1000 replicates, I generated a landscape with
100 sites by sampling exactly from a probability distribution defined by the
interaction coefficients in that figure (Appendix A). Each of the methods
described below was then evaluated on its ability to correctly infer that the
two shrub species competed with one another, despite their frequent
co-occurrence.

I also simulated a second set of landscapes using a stochastic community model
based on generalized Lotka-Volterra dynamics, as described in the Appendix. In
these simulations, each species pair was randomly assigned to either compete for
a portion of the available carrying capacity (negative interaction) or to act as
mutualists (positive interaction).  Here, mutualisms operate by mitigating the
effects of intraspecific competition on each partner's death rate.  For these
analyses, I simulated landscapes with up to 20 species and 25, 200, or 1600
sites (50 replicates per landscape size; see Appendix).

#### Recovering species interactions from simulated data.
I used Markov networks and several other techniques for determining the sign and
strength of the associations between pairs of species (Appendix B).

@@

For the Markov networks, I used the rosalia package [@harris_rosalia_2015],
as described above. For the analyses with 20 species, I added a very weak
logistic prior distribution on the $\alpha$ and $\beta$ terms with scale 2 to
ensure that the model estimates were always finite. The bias introduced by this
prior should be small: the 95% credible interval on $\beta$ only requires that
one species' effect on the odds of observing a different species to be less than
a factor of 1500 (which is not much of a constraint). The logistic distribution
was chosen because it is convex and has a similar shape to the Laplace
distribution used in LASSO regularization (especially in the tails), but unlike
the Laplace distribution it is differentiable everywhere and does not force any
estimates to be exactly zero. To confirm that this procedure produced reliable
estimates, I evaluated it on 50 bootstrap replicates (Appendix).

I also evaluated five alternative methods from the existing literature (plus a
novel variant of the fifth method). The first two alternative interaction
measures were the sample correlations and the partial correlations between each
pair of species' data vectors on the landscape [@albrecht_spatial_2001;
@faisal_inferring_2010]. Because partial correlations are undefined for
landscapes with perfectly-correlated species pairs, I used a regularized
estimate based on James-Stein shrinkage, as implemented in the corpcor package’s
`pcor.shrink` function with the default settings [@schafer_corpcor_2014]. In the
context of non-Gaussian data, the partial correlation can be thought of as a
computationally efficient approximation to the full Markov network model
[@loh_structure_2013]. This sort of model is very common for estimating
relationships among genes and gene products [@friedman_sparse_2008].

@@

The last two estimators used the latent correlation matrix estimated by the
BayesComm package [@golding_bayescomm_2015] in order to evaluate the untested
claim from recent papers that the correlation coefficients estimated by "joint
species distribution models" provide an accurate assessment of species’ pairwise
interactions [@pollock_understanding_2014; see also @harris_generating_2015].  
In addition to using the posterior mean of the correlation coefficient
[@pollock_understanding_2014], I also used the posterior mean of the *partial*
correlation, which might provide a better estimate of species' conditional
associations.

#### Evaluating model performance.
For the simulated landscapes based on Figure 1, method evaluation was fairly
qualitative: any method whose test statistic for the two shrubs indicated a
negative relationship passed; other methods failed (Appendix). For the null
model (Pairs), I calculated statistical significance using its $Z$-score. For
the Markov network, I used the Hessian matrix to generate approximate confidence
intervals and noted whether these intervals included zero.

For the larger landscapes, I evaluated the relationship between each method's
estimates and the "true" interaction strengths among all of the species pairs
from the simulated landscapes. This kind of evaluation is important because
ecologists need a consistent way to know how strong species interactions are
regardless of which species were present in a particular data set or how many
observations were taken. Because the different methods mostly describe species
interactions on different scales (e.g. correlations versus $Z$ scores versus
regression coefficients), I used linear regression through the origin to rescale
the different estimates produced by each method so that they had a consistent
interpretation. After rescaling each method’s estimates, I calculated squared
errors between the scaled interaction estimates and “true” interaction values
across all the simulated data sets. These squared errors determined the
proportion of variance explained for different combinations of method and
landscape size (compared with a baseline model that assumed all interaction
strengths to be zero).

@@

#### Three species.
As shown in Figure 1, the marginal relationship between the two shrub species
was positive---despite their competition for space at a mechanistic level---due
to indirect effects of the dominant tree species. As a result, the correlation
between these species was positive in 94% of replicates, and the
randomization-based null model falsely reported positive associations 100% of
the time. Worse, more than 98% of false conclusions from the null model were
statistically significant. The partial correlation and Markov network estimates,
on the other hand, each correctly isolated the direct negative interaction
between the shrubs from their positive indirect interaction 94% of the time
(although the confidence intervals overlapped zero in most replicates).

#### Twenty species.
The accuracy of the evaluated methods varied substantially, depending on
the parameters that produced the simulated communities (Figure 3A). In general,
however, there was a consistent ordering, with the four methods that controlled
for indirect effects of species interactions performing the best: the Markov
network explained the largest portion of the variance (35% overall), followed
by the generalized linear models (30%), partial correlations from the raw
presence-absence data (28%), and partial correlations from BayesComm, the
joint species distribution model (26%).  The benefit of choosing the full
Markov network over the other three methods was largest on the smaller
landscapes, which are also the ones that are most representative of typical
analyses in this field [@gotelli_empirical_2009].

The three methods that did not attempt to control for indirect interactions all
explained less than 20% of the variance. Of these, the sample correlation matrix
based on the raw data performed the best (19%), followed by the null model
(15%) and BayesComm's correlation matrix (11%). Although these last three
methods had different $R^2$ values, there was a close mapping among
their estimates (after controlling for the size of the simulated landscapes;
Figure 3B).  This suggests that the correlation matrices from joint species
distribution models and especially the effect sizes from null models only
contain noisy versions of the same information that could be obtained more
easily and interpretably by calculating correlation coefficients between
species' presence-absence vectors.

The bootstrapped results confirmed that the rosalia package's estimates of
species' conditional relationships were reasonably insensitive to sampling
variation (Appendix), and that the above ranking of the different methods was
robust (Appendix).

@@

Despite their limitations, Markov networks have enormous potential to improve
ecological understanding. In particular, they are less vulnerable than some
of the most commonly-used methods to mistakenly identifying positive species
interactions between competing species. They also provide a simple answer to the
question of how competition should affect a species' overall prevalence, which
was a major flashpoint for the null model debates in the 1980’s
[@roughgarden_competition_1983; @strong_ecological_1984]. Equation 1 can be used
to calculate the expected prevalence of a species in the absence of biotic
influences [$\frac{1}{1 + e^{-\alpha}}$; @lee_learning_2012]. Competition's
effect on prevalence in a Markov network can then be calculated by subtracting
this value and the observed prevalence (cf Figure 2D). This kind of insight
would have been difficult to obtain without a generative model that makes
predictions about the consequences of species interactions; null models (which
presume *a priori* that interactions do not exist) have no way to make such
predictions.

Markov networks---particularly the Ising model for binary networks---have been
studied in statistical physics for nearly a century [@cipra_introduction_1987],
and the models' properties, capabilities, and limits are well-understood in a
huge range of applications. Using the same framework for species interactions
would thus allow ecologists to tap into an enormous set of existing
discoveries and techniques for dealing with indirect effects, stability, and
alternative stable states. Numerous other extensions are possible: for example,
Markov networks can be fit with a mix of discrete and continuous variables,
and likely states of the interaction network can be modeled as a function of
the local abiotic environment [@lee_learning_2012], which would provide a
rigorous and straightforward way to incorporate whole networks of biotic
interactions into species distribution models. Estimating each species'
conditional relationship with both biotic and abiotic factors in the same model
should lead to a better understanding of both kinds of ecological process. There
are even methods [@whittam_species_1981; @tjelmeland_markov_1998] that would
allow one species to affect the sign or strength of the relationship between
two other species, tipping the balance between facilitation and exploitation
[@bruno_inclusion_2003].

Finally, the results presented here have important implications for
ecologists' continued use of null models to draw inferences about species
interactions. Null and neutral models can be very useful for clarifying our
thinking about the numerical consequences of species' richness and abundance
patterns [@harris_occupancy_2011; @xiao_strong_2015], but deviations from a
particular null model must be interpreted with care
[@roughgarden_competition_1983]. Even in small networks with three species, it
may simply not be possible to implicate individual species pairs or specific
ecological processes like competition by rejecting a general-purpose null
[@gotelli_empirical_2009], especially when the test statistic is effectively
just a correlation coefficient (Figure 3B). Simultaneous estimation of multiple
ecological parameters seems like a much more promising approach: to the extent
that the models' relative performance on real data sets is similar to the range
of results shown in Figure 3A, scientists in this field could often double their
explanatory power by switching from null models to Markov networks (or increase
it by nearly as much with linear or generalized linear approximations).
Regardless of the specific methods ecologists ultimately choose, the most
important consideration is clearly that ecologists need to control for indirect
effects and estimate conditional relationships between species, rather than
marginal ones.

@@

**Figure 2.**
**A.** A small Markov network with two species.  The abiotic environment favors
the occurrence of both species ($\alpha >0$), particularly species 2
($\alpha_2 > \alpha_1$). The  negative $\beta$ coefficient linking these two
species implies that they co-occur less than expected under independence.
**B.** Relative probabilities of all four possible presence-absence combinations
for Species 1 and Species 2. The exponent includes $\alpha_1$ whenever Species 1
is present ($y_1 = 1$), but not when it is absent ($y_1 = 0$).  Similarly, the
exponent includes $\alpha_2$ only when species $2$ is present ($y_2 = 1$), and
$\beta$ only when both are present ($y_1y_2 = 1$). The normalizing constant $Z$,
ensures that the four relative probabilities sum to 1.  In this case, $Z$ is
about 18.5.  **C.** Using the probabilities, we can find the expected
frequencies of all possible co-occurrence patterns between the two species of
interest. **D.** If $\beta$ equaled zero (e.g. if the species no longer competed
for the same resources), then the reduction in competition would allow each
species to increase its occurrence rate and the co-occurrence deficit would
be eliminated.

**Figure 3.**
**A.** Proportion of variance in interaction coefficients explained by each
method across varying numbers of sampled locations.
**B.** The $Z$-scores produced by the null model (i.e. the Pairs software) for
each pair of species are very predictable from the correlation between the
presence-absence vectors of those same species and from the number of sites on
the landscape.

>>>>

\setlength{\parskip}{0pt}

@@

If nontrophic species interactions, such as competition, are important drivers
of community assembly, then ecologists might expect to see their influence in
species composition data [@macarthur_population_1958; @diamond_island_1975].
Despite decades of work and several major controversies, however
[@lewin_santa_1983; @strong_ecological_1984; @gotelli_swap_2003;
@connor_checkered_2013], existing methods for detecting competition's effects
on community structure are unreliable [@gotelli_empirical_2009]. In particular,
species' effects on one another can become lost in the complex web of direct and
indirect interactions in real assemblages. For example, the competitive
interaction between the two shrub species in Figure 1A can become obscured by
their shared tendency to occur in unshaded areas (Figure 1B). While ecologists
have long known that indirect effects can overwhelm direct ones at the landscape
level [@dodson_complementary_1970; @levine_competitive_1976], the vast majority of our methods for drawing
inferenes from observational data do not control for these effects
[e.g. @diamond_island_1975; @strong_ecological_1984; @gotelli_empirical_2009;
@veech_probabilistic_2013; @pollock_understanding_2014]. To the extent that
indirect interactions like those in Figure 1 are generally important
[@dodson_complementary_1970], existing methods will thus not generally provide
much evidence regarding species' direct effects on one another. The goal of this
paper is to resolve this long-standing problem.

@@

Following recent work by @azaele_inferring_2010 and @fort_statistical_2013, this
paper shows that Markov networks [undirected graphical models also known as
Markov random fields; @murphy_machine_2012] can provide a framework for
understanding the landscape-level consequences of pairwise species interactions,
and for detecting them from observed presence-absence matrices. Markov networks
have been used in many scientific fields for decades in similar contexts, from
physics [where nearby particles interact magnetically; @cipra_introduction_1987]
to spatial statistics [where adjacent grid cells have correlated values;
@harris_contact_1974; @gelfand_modelling_2005]. While community ecologists
explored some related approaches in the 1980's [@whittam_species_1981],
they used severe approximations that led to unintelligible results
[e.g. "probabilities" greater than one; @gilpin_factors_1982].

Below, I introduce Markov networks and show how they can be used to simulate
landscape-level data or to make exact predictions about the direct and
indirect consequences of possible interaction matrices. Then, using simulated
data sets where the "true" interactions are known, I compare this approach with
several existing methods. Finally, I discuss opportunities for extending the
approach presented here to other problems in community ecology, e.g. quantifying
the overall effect of species interactions on occurrence rates
[@roughgarden_competition_1983] and disentangling the effects of biotic versus
abiotic interactions on species composition [@kissling_towards_2012;
@pollock_understanding_2014].

@@

#### Markov networks.
Markov networks provide a framework for translating back and forth between the
conditional relationships among species (Figure 1C) and the kinds of species
assemblages that these relationships produce. Here, I show how a set of
conditional relationships can be used to determine the how groups of species
can co-occur. Methods for estimating conditional relationships from data are
discussed in the next section.

@@

Here, $\alpha_{i}$ is an intercept term determining the amount that the presence
of species $i$ contributes to the log-probability of $\vec{y}$; it directly
controls the prevalence of species $i$. Similarly, $\beta_{ij}$ is the amount
that the co-occurrence of species $i$ and species $j$ contributes to the
log-probability; it controls the conditional relationship between two species,
i.e. the probability that they will be found together, after controlling for the
other species in the network (Figure 2A, Figure 2B). For example, $\beta_{ij}$
might have a value of $+2$ for two mutualists, indicating that the odds of
observing one species are $e^2$ times higher in sites where its partner is
present than in comparable sites where its partner is absent. Because the
relative probability of a presence-absence vector increases when
positively-associated species co-occur and decreases when negatively-associated
species co-occur, the model tends---all else equal---to produce assemblages that
have many pairs of positively-associated species and relatively few pairs of
negatively-associated species (exactly as an ecologist might expect).

@@

#### Estimating $\alpha$ and $\beta$ coefficients from presence-absence data.
In the previous section, the values of $\alpha$ and $\beta$ were known and the
goal was to make predictions about possible species assemblages. In
practice, however, ecologists will often need to estimate the parameters from
an observed co-occurrence matrix (i.e. from a matrix of ones and zeros
indicating which species are present at which sites). When the number of species
is reasonably small, one can compute exact maximum likelihood estimates for all
of the $\alpha$ and $\beta$ coefficients given a presence-absence matrix by
optimizing $p(\vec{y}; \alpha, \beta)$. Fully-observed Markov networks like the
ones considered here have unimodal likelihood surfaces [@murphy_machine_2012],
ensuring that this procedure will always converge on the global maximum. This
maximum represents the unique combination of $\alpha$ and $\beta$ coefficients
that would be expected to produce exactly the observed co-occurrence frequencies
on average [i.e. it matches the sufficient statistics of the model distribution
to the sufficient statistics of the data; @murphy_machine_2012]. I used the
rosalia package [@harris_rosalia_2015] for the R programming language
[@r_core_team_r_2015] to optimize the Markov network parameters. The package was
named after Santa Rosalia, the patron saint of biodiversity, whose supposedly
miraculous healing powers played an important rhetorical role in the null model
debates of the 1970's and 1980's [@lewin_santa_1983].

#### Simulated landscapes.
In order to compare different methods, I simulated two sets of landscapes using
known parameters. The first set included the three competing species shown in
Figure 1. For each of 1000 replicates, I generated a landscape with 100 sites by
sampling from a probability distribution defined by the figure's interaction
coefficients (Appendix 1). Each of the methods described below was then
evaluated on its ability to correctly infer that the two shrub species competed
with one another, despite their frequent co-occurrence.

I also simulated a second set of landscapes using a stochastic community model
based on generalized Lotka-Volterra dynamics, as described in Appendix 2. In
these simulations, each species pair was randomly assigned to either compete for
a portion of the available carrying capacity (negative interaction) or to act as
mutualists (positive interaction).  Here, mutualisms operate by mitigating the
effects of intraspecific competition on each partner's death rate.  For these
analyses, I simulated landscapes with up to 20 species and 25, 200, or 1600
sites (50 replicates per landscape size; see Appendix 2).

#### Recovering species interactions from simulated data.
I compared seven techniques for determining the sign and strength of the
associations between pairs of species from simulated data (Appendix 3).

@@

First, I used the rosalia package [@harris_rosalia_2015] to fit Markov newtork
models,  as described above. For the analyses with 20 species, I added a very
weak logistic prior distribution on the $\alpha$ and $\beta$ terms with scale 2
to ensure that the model estimates were always finite. The bias introduced by
this prior should be small: the 95% credible interval on $\beta$ only requires
that one species' effect on the odds of observing a different species to be less
than a factor of 1500 (which is not much of a constraint). The logistic
distribution was chosen because it is convex and has a similar shape to the
Laplace distribution used in LASSO regularization (especially in the tails), but
unlike the Laplace distribution it is differentiable everywhere and does not
force any estimates to be exactly zero. To confirm that this procedure produced
stable estimates, I compared its estimates on 50 bootstrap replicates
(Appendix 4).

I also evaluated five alternative methods from the existing literature (plus a
novel variant of the fifth method). The first two alternative interaction
measures were the sample correlations and the partial correlations between each
pair of species' data vectors on the landscape [@albrecht_spatial_2001;
@faisal_inferring_2010]. In the context of non-Gaussian data, the partial
correlation can be thought of as a computationally efficient approximation to
the full Markov network model [@loh_structure_2013]. This sort of model is very
common for estimating relationships among genes and gene products
[@friedman_sparse_2008]. Because partial correlations are undefined for
landscapes with perfectly-correlated species pairs, I used a regularized
estimate based on James-Stein shrinkage, as implemented in the corpcor package’s
`pcor.shrink` function with the default settings [@schafer_corpcor_2014].

@@

The last two estimators used the latent correlation matrix estimated by the
BayesComm package [@golding_bayescomm_2015] in order to evaluate the recent
claim that the correlation coefficients estimated by "joint species distribution
models" provide an accurate assessment of species’ pairwise interactions
[@pollock_understanding_2014; see also @harris_generating_2015]. In addition to
using the posterior mean correlation [@pollock_understanding_2014], I also used
the posterior mean *partial* correlation, which might be able to control for
indirect effects.

#### Evaluating model performance.
For the simulated landscapes based on Figure 1, I assessed whether each method's
test statistic indicated a positive or negative relationship between the two
shrubs (Appendix 1). For the null model (Pairs), I calculated statistical
significance using its $Z$-score. For the Markov network, I used the Hessian
matrix to generate approximate confidence intervals and noted whether these
intervals included zero.

I then evaluated the relationship between each method's estimates and the "true"
interaction strengths among all of the species pairs from the larger simulated
landscapes. This determined which the methods provide a consistent way to know
how strong species interactions are---regardless of which species were present
in a particular data set or how many observations were taken. Because the
different methods mostly describe species interactions on different scales
(e.g. correlations versus $Z$ scores versus regression coefficients), I used
linear regression through the origin to rescale the different estimates produced
by each method so that they had a consistent interpretation. After rescaling
each method’s estimates, I calculated squared errors between the scaled
interaction estimates and “true” interaction values across all the simulated
data sets. These squared errors determined the proportion of variance explained
for different combinations of method and landscape size (compared with a
null model that assumed all interaction strengths to be zero).

@@

#### Three species.
As shown in Figure 1, the marginal relationship between the two shrub species
was positive---despite their competition for space at a mechanistic level---due
to indirect effects of the dominant tree species. As a result, the correlation
between these species was positive in 94% of replicates, and the
randomization-based null model falsely reported positive associations 100% of
the time. Worse, more than 98% of these false conclusions were statistically
significant. The partial correlation and Markov network estimates, on the other
hand, each correctly isolated the direct negative interaction between the shrubs
from their positive indirect interaction 94% of the time (although the
confidence intervals overlapped zero in most replicates).

#### Twenty species.
Despite some variability across contexts (Figure 3A), the four methods that
controlled for indirect effects clearly performed the best: the Markov network
explained the largest portion of the variance in the "true" interaction
coefficients (35% overall), followed by the generalized linear models (30%),
partial correlations from the raw presence-absence data (28%), and partial
correlations from BayesComm, the joint species distribution model (26%).  The
benefit of choosing the full Markov network over the other three methods was
largest on the smaller landscapes, which are also the ones that are most
representative of typical analyses in this field [@gotelli_empirical_2009].

The three methods that did not attempt to control for indirect interactions all
explained less than 20% of the variance. Of these, the sample correlation matrix
based on the raw data performed the best (19%), followed by the null model
(15%) and BayesComm's correlation matrix (11%). Although these last three
methods had different $R^2$ values, there was a close mapping among
their estimates (especially after controlling for the size of the simulated
landscapes; Figure 3B).  This suggests that the effect sizes from the null model
(and, to a lesser extent, the correlation matrices from joint species
distribution models) only contain noisy versions of the same information that
could be obtained more easily and interpretably by calculating correlation
coefficients between species' presence-absence vectors.

Bootstrap resampling indicated that the above ranking of the different methods
was robust (Appendix 3). In particular, the 95% confidence interval of the
bootstrap distribution indicated that the Markov network's overall $R^2$ value
was between 14 and 18 percent higher than the second-most effective
method (generalized linear models) and between 2.12 and 2.38 times higher than
could be achieved by the null model (Pairs).  Bootstrap resampling of a 200-site
landscape also confirmed that the rosalia package's estimates of species'
conditional relationships were reasonably robust to sampling variation
(Appendix 4).

@@

Despite their limitations, Markov networks have enormous potential to improve
ecological understanding. In particular, they are less vulnerable than some
of the most commonly-used methods to mistakenly identifying positive species
interactions between competing species, and can make precise statements about
the conditions whre indirect interactions will overwhelm direct ones. They also
provide a simple answer to the question of how competition should affect a
species' overall prevalence, which was a major flashpoint for the null model
debates in the 1980’s [@roughgarden_competition_1983; @strong_ecological_1984].
Equation 1 can be used to calculate the expected prevalence of a species in the
absence of biotic influences [$\frac{1}{1 + e^{-\alpha}}$; @lee_learning_2012].
Competition's effect on prevalence in a Markov network can then be calculated by
subtracting this value from the observed prevalence (cf Figure 2D). This kind of
insight would have been difficult to obtain without a generative model that
makes predictions about the consequences of species interactions; null models
(which presume *a priori* that interactions do not exist) have no way to make
such predictions.

Markov networks---particularly the Ising model for binary networks---have been
studied in statistical physics for nearly a century [@cipra_introduction_1987],
and the models' properties, capabilities, and limits are well-understood in a
huge range of applications. Using the same framework for species interactions
would thus allow ecologists to tap into an enormous set of existing
discoveries and techniques for dealing with indirect effects, stability, and
alternative stable states. Numerous other extensions are possible: for example,
the states of the interaction network can be modeled as a function of
the local abiotic environment [@lee_learning_2012], which would provide a
rigorous and straightforward approach to the difficult and important task of
incorporating whole networks of biotic interactions into species distribution
models [@kissling_towards_2012; @pollock_understanding_2014], leading to a
better understanding of the interplay between biotic and abiotic effects on
community structure. There are even methods [@whittam_species_1981;
@tjelmeland_markov_1998] that would allow one species to affect the sign or
strength of the relationship between two other species, tipping the balance
between facilitation and exploitation [@bruno_inclusion_2003].

Finally, the results presented here have important implications for
ecologists' continued use of null models for studying species interactions. Null
and neutral models can be useful for clarifying our thinking about the numerical
consequences of species' richness and abundance patterns
[@harris_occupancy_2011; @xiao_strong_2015], but deviations from a particular
null model must be interpreted with care [@roughgarden_competition_1983]. Even
in small networks with three species, it may simply not be possible to implicate
individual species pairs or specific ecological processes like competition by
rejecting a general-purpose null [@gotelli_empirical_2009], especially when the
test statistic is effectively just a correlation coefficient (Figure 3B).
Simultaneous estimation of multiple ecological parameters seems like a much more
promising approach: to the extent that the models' relative performance on real
data sets is similar to the range of results shown in Figure 3A, scientists in
this field could often double their explanatory power by switching from null
models to Markov networks (or increase it substantially by with linear or
generalized linear approximations). Regardless of the methods ecologists
ultimately choose, controlling for indirect effects could clearly improve our
understanding of species' direct effects on one another and on community
structure.

@@

**Figure 2.**
**A.** A small Markov network with two species.  The abiotic environment favors
the occurrence of both species ($\alpha >0$), particularly species 2
($\alpha_2 > \alpha_1$). The  negative $\beta$ coefficient linking these two
species implies that they co-occur less than expected under independence.
**B.** Relative probabilities of all four possible presence-absence combinations
for Species 1 and Species 2. The exponent includes $\alpha_1$ whenever Species 1
is present ($y_1 = 1$), but not when it is absent ($y_1 = 0$).  Similarly, the
exponent includes $\alpha_2$ only when species $2$ is present ($y_2 = 1$), and
$\beta$ only when both are present ($y_1y_2 = 1$). The normalizing constant $Z$,
ensures that the four relative probabilities sum to 1.  In this case, $Z$ is
about 18.5.  **C.** We can find the expected frequencies of all possible
co-occurrence patterns between the two species of interest. **D.** If
$\beta_{12}$ equaled zero (e.g. if the species no longer competed for the same
resources), then the reduction in competition would allow each species to
increase its occurrence rate and the co-occurrence deficit would be eliminated.

**Figure 3.**
**A.** Proportion of variance in interaction coefficients explained by each
versus number of sampled locations.
**B.** The $Z$-scores produced by the null model ("Pairs") for each pair of
species are very predictable from the correlation between the presence-absence
vectors of those same species and from the number of sites on the landscape.

\setlength{\parskip}{0pt}
\setlength{\parindent}{-1em}
\setlength{\leftskip}{1em}
